---
title: "About"
editor_options: 
  chunk_output_type: console
execute: 
  eval: false
engine: knitr
shift-heading-level-by: 1
---

# Working with .janno context information in R

For data analysis and visualization an R-based workflow may be helpful. This is especially relevant if we want to incorporate spatial and temporal information.

## Adding new information to the .janno file

As a first step we add the new information we received from Prof. P. to our Poseidon package `ice`. We do so by editing the .janno file at `ice/ice.janno`.

Starting from

::: {.table-responsive}

|Poseidon_ID|Genetic_Sex|Group_Name|Latitude|Longitude|
|-----------|-----------|----------|--------|---------|
|ice        |M          |iceG      |46.77   |10.83    |

:::

we add three additional columns as defined in the documentation here.

::: {.table-responsive}

|Date_Type  |Date_C14_Uncal_BP|Date_C14_Uncal_BP_Err|
|-----------|-----------------|---------------------|
|C14        |4555             |34                   |

:::

As we want to work particularly neatly we document this change to the package in a changelog, which we add to the package in the form of a `CHANGELOG.md` file.

```{filename="CHANGELOG.md"}
- V 0.2.0: Added a radiocarbon date for the ice sample.
```

We also update the package version number in our `POSEIDON.yml` file and link the changelog file there:

```{filename="POSEIDON.yml"}
poseidonVersion: 2.7.1
title: ice
packageVersion: 0.2.0
lastModified: 2024-01-24
...
changelogFile: CHANGELOG.md
```

After implementing the change we should again check the structural validity of the package with `trident validate`.

```{bash}
trident validate -d ice
```

## Loading .janno files in R

We previously prepared the 

```{r}
pca_out <- readRDS("scratch/pca_out2.rds")
```

```{r}
library(magrittr)
library(ggplot2)

my_janno <- janno::read_janno("scratch/data/2015_AllentoftNature-2.1.1/2015_AllentoftNature.janno")

my_janno_age <- my_janno %>% janno::process_age()

my_janno_age %>%
  dplyr::select(Poseidon_ID, Date_BC_AD_Median, Date_BC_AD_Sample) %>%
  dplyr::arrange(Date_BC_AD_Median) %>%
  dplyr::mutate(Poseidon_ID = factor(Poseidon_ID, levels = Poseidon_ID)) %>%
  tidyr::unnest(cols = Date_BC_AD_Sample) %>%
  ggplot() +
  geom_violin(aes(x = Poseidon_ID, y = Date_BC_AD_Sample))
```





